\Large{All in the family:\\Anna and Lisa Hahner's finishing times in\\the 2016 women's Olympic marathon}

### 2017-02-01 paper/paper.tex
a few changes

On August 14, 2016, at 9:30 in the morning, the Women's Olympic
marathon kicked off in Rio de Janeiro, Brazil, when 156 runners from
80 countries across the world took off from the starting line en route
to their destination 42.195 kilometers away. Two hours, twenty-four
minutes, and four seconds later, Jemima Sumgong of Kenya would be the
first to cross the finishline and take home gold; Sumgong was just 3.5
minutes behind her personal best time in the marathon. Approximately
21 minutes later, twin marathoners from Germany, Anna and Lisa Hahner,
would cross the finishline together, holding hands and celebrating a
personal victory. Although the Hahners would finish 81st and 82nd,
respectively, with times slightly more than 18 minutes slower than
corresponding personal bests, Anna Hahner would describe their joint
finish as a ``magical moment.''
\href{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}
The media quickly picked up on the Hanher story as an image of the
beaming twins finishing hand-in-hand captured a public audience. Many
believed the moment was a reflection of the Olympic spirit.

Not everyone agreed with this rosy interpretation. The twins' happy
facial expressions at the finish were portrayed as a bit contrived
(smiling like ``Honigkuchenpferde''---a Honigkuchenpferd is a German
cookie in the shape of a horse), and the sports director of the German
Athletics Federation, Thomas Kurschilgen, stirred up controversy when
he suggested that the Hahners' photo-finish was no coincidence.
Kurschilgen averred that the twins slowed down so as to finish
simultaneously and sought to create a spectacle and ``generate media
attention.'' Kurschilgen justified his charge with the fact that the
twins were at least 18 minutes behind their personal best
times.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
Not surprisingly, these accusations were denied by the Hahners, who
claimed that their simultaneous finish was simply an unintended
coincidence.

@@

What happened in the women's Olympic marathon, and how might we assess
whether the Hahner finish was coincidental or intentional? These two
interpretations are clearly at odds. If the former, then the Hahner
twins are to be celebrated and their finish treated as an expression
of the spirit behind the Olympic games. If the latter, though, then
the Hahner twins may have violated this spirt by not trying hard
enough. It is perhaps too easy for us to write such a glib
sentence---neither of us cannot fathom being able to complete a
marathon anywhere in the vicinity of two and a half hours---but we
nonetheless want to know what the data from the Olympic marathon data
tell us. Was the women's Olympic marathon a lovely coincidence or
something else?

Among female Olympic marathoners, the Hahner twins were not alone in
their familial ties, and we draw on this in the analysis that follows.
The marathon also featured twins from North Korea, Kim Hye-song and
Kim Hye-gyong, who had identical times and finished 10th and 11th,
respectively. The Kim finish, unlike the Hahner finish, appears devoid
of post-race controversy. Moreover, three triplets from Estonia
competed in the Rio marathon, although only two, Lily Luik and Leila
Luik, finished it, in 97th and 114th place, respectively. The third
Estonia triplet, Liina Luik, recorded what is known as a DNF---an
abbreviation that means did not finish, a term that we will use
throughout this article.

\section*{Marathon data}

For each participant who started the women's Olympic marathon, we know
several things: personal best marathon time prior to the 2016 Olympic
games; split times from the Rio marathon course at 5 kilometers, 10
kilometers, and so forth; and, finishing time. We cannot directly
observe the effort that an individual put into the race, and we do not
know why some runners have DNF results; some runners may have injured
themselves on the course and accordingly dropped out, and others may
have dropped out, uninjured, in anticipation of an unsatisfactory
result. Of the 156 marathon starters, 133 completed the race and 23
DNFed at various locations throughout the course. The overall DNF rate
was thus $\frac{23}{156} \approx 0.15$, and the relatively small
sample size at our disposal---156 runners---means that a 95\%
confidence for this rate is fairly wide, namely,
$\left(0.098, 0.22\right)$.

Key in the analysis that follows is the deviation between a runner's
Rio time and her prior personal best time in the marathon. Of course
there is variance across runners in these deviations, and we want to
know if patterns in the deviations support an honest effort by the
Hanher twins or perhaps something else.

As a brief overview of the race, Figure \ref{fig:secondsbehind}
describes each runner's status at the various split times on the
marathon course.  Each dot in the figure depicts a recorded split and
the number of seconds each runner was behind the race leader at the
time.  There are more dots at earlier splits due to the accumulation
of DNFs.  Regarding twins and triplets, the Estonian Luik triplets are
denoted with blue dots; there are three blue dots through the 20
kilometer split, but the third triplet did not record a half-marathon
split.  The orange dots in the figure represent the Hahner twins; Lisa
Hahner was ahead of her sister through 15 kilometers at which point
Anna surged ahead.  There are two red dots representing the North
Korean Kim twins although this is not visually apparent; the Kim twins
had identical split times during the marathon.

>>>>

On August 14, 2016, at 9:30 in the morning, the Women's Olympic
marathon kicked off in Rio de Janeiro, Brazil, when 156 runners from
80 countries across the world left the starting line en route to their
destination 42.195 kilometers away. Two hours, twenty-four minutes,
and four seconds later, Jemima Sumgong of Kenya would be the first to
cross the finishline and take home gold; Sumgong was just 3.5 minutes
behind her personal best time in the marathon. Approximately 21
minutes later, twin marathoners from Germany, Anna and Lisa Hahner,
would cross the finishline together, holding hands and celebrating a
personal victory. Although the Hahners would finish 81st and 82nd,
respectively, with times slightly more than 18 minutes slower than
corresponding personal bests, Anna Hahner would describe their joint
finish as a ``magical moment.''
\href{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}
The media quickly picked up on the Hanher story as an image of the
beaming twins finishing hand-in-hand captured a public audience. Many
believed the moment was a reflection of the Olympic spirit.

Not everyone agreed with this rosy interpretation. The twins' happy
facial expressions at the finish were portrayed as a bit contrived
(smiling like ``Honigkuchenpferde''---a Honigkuchenpferd is a German
cookie in the shape of a horse), and the sports director of the German
Athletics Federation, Thomas Kurschilgen, stirred up controversy when
he suggested that the Hahners' photo-finish was no coincidence.
Kurschilgen averred that the twins slowed down so as to finish
simultaneously and create a spectacle which would ``generate media
attention.'' Kurschilgen justified his charge with the fact that the
twins ran in the Rio marathon at least 18 minutes slower their
personal best times prior to the
Olympics.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
Not surprisingly, these accusations were denied by the Hahner twins,
who claimed that their simultaneous finish was simply an unintended
coincidence.

@@

What happened in the women's Olympic marathon, and how might we
develop a statistical approach that assesses whether the Hahner finish
was coincidental or intentional?  These two interpretations are
clearly at odds. If the former, then the Hahner twins are to be
celebrated and their finish treated as an expression of the spirit
behind the Olympic games. If the latter, though, then the twins may
have violated this spirt by not trying hard enough. It is perhaps too
easy for us to write such a glib sentence---neither of us cannot
fathom being able to complete a marathon anywhere in the vicinity of
two and a half hours---but we nonetheless want to know what the data
from the Olympic marathon data tell us.  Was the Hahner finish in the
2016 women's Olympic marathon a lovely coincidence or something else?

Among female Olympic marathoners, the Hahner twins were not alone in
their familial ties, and we draw on this in the analysis that follows.
The marathon also featured twins from North Korea, Kim Hye-song and
Kim Hye-gyong, who posted identical times and finished 10th and 11th
in the race, respectively. The Kim finish, unlike the Hahner finish,
appears devoid of post-race controversy. Moreover, three triplets from
Estonia competed in the Rio marathon, although only two, Lily Luik and
Leila Luik, finished it, in 97th and 114th place, respectively. The
third Estonia triplet, Liina Luik, recorded what is known as a
DNF---an abbreviation that means did not finish, a term that we will
use throughout this article.

\section*{Marathon data and our research design}

For each participant who started the women's Olympic marathon, we know
several things: personal best marathon time prior to the 2016 Olympic
games; split times from the Rio marathon course at 5 kilometers, 10
kilometers, and so forth; and, finishing time. We cannot directly
observe the effort that an individual put into the race, and we do not
know why some runners have DNF results.  Some runners may have injured
themselves on the course and accordingly dropped out, and others may
have dropped out, uninjured, in anticipation of an unsatisfactory
result. Of the 156 marathon starters, 133 completed the race and 23
DNFed at various locations throughout the course. The overall DNF rate
was thus $\frac{23}{156} \approx 0.15$, and the relatively small
sample size at our disposal means that a 95\% confidence for this rate
is fairly wide, namely, $\left(0.098, 0.22\right)$.  Key in the
analysis that follows is the deviation between a runner's Rio time and
her prior personal best time in the marathon.  While there is variance
across runners in these deviations, they nonetheless may be able to
provide us with insight as to the Hahner twins' performance in the
marathon.

Our research design has two components.  In the first, we present some
visualizations that present the Hahner twin's results in context; our
interest here is primarily the two Hanhers although we comment on the
Kim twins and Luik triplets when appropriate.  Our visualizations
suggest blah blah bhah.  We then turn to a modeling approach and in
particular a simulation.  In simulated Olympic marathons, which we
develop based on relationships (excluding twins and triplets) between
known personal best times and observed Rio finishing times, we find
that the Hahner twins finished suspiciously close to each other given
the disparty between their personal best times.  OR THEY DID NOT.

\section*{Visualizing the Olympic marathon}

SOME PLOTS HERE.  Maybe PB versus FINAL and HALF vs Final.

As a brief overview of the race, Figure \ref{fig:secondsbehind}
describes each runner's status at the various split times on the
marathon course.  Each dot in the figure depicts a recorded split and
the number of seconds each runner was behind the race leader at the
time.  There are more dots at earlier splits due to the accumulation
of DNFs.  Regarding twins and triplets, the Estonian Luik triplets are
denoted with blue dots; there are three blue dots through the 20
kilometer split, but the third triplet did not record a half-marathon
split.  The orange dots in the figure represent the Hahner twins; as
noted in the figure, Lisa Hahner was ahead of her sister through 15
kilometers at which point Anna surged ahead.  Lastly, Figure
\ref{fig:secondsbehind} contains two red dots representing the North
Korean Kim twins; this is not visually apparent because the Kim twins
had identical split times during the marathon.
